---
title: "Comparison of performances on our data"
output: 
  workflowr::wflow_html:
    toc: TRUE
    toc_float: TRUE
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


We compare the performance of peco with Seurat, Cyclone, Oscope and reCAT using our data in six-fold cross-validation. Here we provide the code that we used to predict continuous cel cycle phase in each method and to compute prediction error between FUCCI phase and predicted cell cycle phase. 



## Set up

Load packages 

```{r, message=FALSE}
library(SingleCellExperiment)
library(peco)
```

Prepare training/testing data

```{r, eval = F}
sce <- readRDS("data/sce-final.rds")
sce <- sce[grep("ENSG", rownames(sce)),]
fdata <- data.frame(colData(sce))
fdata <- data.frame(rowData(sce))
counts <- data.frame(assay(sce, "counts"))

sce_normed <- data_transform_quantile(sce)
log2cpm_quant <- assay(sce_normed, "cpm_quantNormed")

inds <- c("NA18511", "NA18855", "NA18870", "NA19098", "NA19101", "NA19160")
theta <- pdata$theta

# function to make training/testing data
makedata_supervised <- function(sce, log2cpm_quant,
                                theta) {
  message("Create data/data_training_test folder \n")
  if (!file.exists("data/data_training_test")) { dir.create("data/data_training_test") }
  library(SingleCellExperiment)
  library(peco)
  pdata <- data.frame(colData(sce))
  fdata <- data.frame(rowData(sce))
  counts <- data.frame(assay(sce))
  counts <- counts[grep("ENSG", rownames(counts)), ]
  log2cpm <- t(log2(1+(10^6)*(t(counts)/pdata$molecules)))

  for (ind in unique(pdata$chip_id)) {
    ii_test <- c(1:nrow(pdata))[which(pdata$chip_id == ind)]
    ii_train <- c(1:nrow(pdata))[which(pdata$chip_id != ind)]

    pdata_test <- pdata[ii_test,]
    pdata_train <- pdata[ii_train,]

    log2cpm_quant_test <- log2cpm_quant[,ii_test]
    log2cpm_quant_train <- log2cpm_quant[,ii_train]
    theta <- pdata$theta
    names(theta) <- rownames(pdata)

    log2cpm_test <- log2cpm[,ii_test]
    log2cpm_train <- log2cpm[,ii_train]

    counts_test <- counts[,ii_test]
    counts_train <- counts[,ii_train]

    theta_test <- theta[ii_test]
    theta_train <- theta[ii_train]

    #sig.genes <- readRDS("output/npreg-trendfilter-quantile.Rmd/out.stats.ordered.sig.476.rds")
    data_training <- list(theta_train=theta_train,
                          log2cpm_quant_train=log2cpm_quant_train,
                          log2cpm_train=log2cpm_train,
                          counts_train=counts_train,
                          pdata_train=pdata_train,
                          fdata=fdata)

    data_test <- list(theta_test=theta_test,
                      log2cpm_quant_test=log2cpm_quant_test,
                      log2cpm_test=log2cpm_test,
                      counts_test = counts_test,
                      pdata_test=pdata_test,
                      fdata=fdata)

    saveRDS(data_training,
            file=file.path(paste0("data/data_training_test/ind_",ind,"_data_training.rds")))
    saveRDS(data_test,
            file=file.path(paste0("data/data_training_test/ind_",ind,"_data_test.rds")))
  }
}
makedata_supervised(sce, log2cpm_quant, theta)
```


## peco

Use peco predictor of 5 genes to predict cell cycle phase. 

```{r, eval = FALSE}
# --- prediction for samples from one individual cell ine
ngenes = 5

for (i in seq_along(inds)) {
  ind = inds[i]
  data_train <- readRDS(paste0("data/data_training_test/ind_", ind, "_data_training.rds"))
  data_test <- readRDS(paste0("data/data_training_test/ind_",ind, "_data_test.rds"))

  fits_all <- readRDS("data/fit.quant.rds")
  genes_all <- names(fits_all)[order(sapply(fits_all,"[[",3), decreasing=T)]

  which_genes <- genes_all[1:ngenes]
  fit_train <- cycle_npreg_insample(
    Y = with(data_train, log2cpm_quant_train[which(rownames(log2cpm_quant_train) %in% which_genes), ]),
    theta = with(data_train, theta_train),
    polyorder=2,
    ncores=4,
    method.trend="trendfilter")

  fit_test <- cycle_npreg_outsample(
    Y_test=with(data_test, log2cpm_quant_test[which(rownames(log2cpm_quant_test) %in% which_genes), ]),
    sigma_est=with(fit_train, sigma_est),
    funs_est=with(fit_train, funs_est),
    method.grid = "uniform",
    method.trend="trendfilter",
    polyorder=2,
    ncores=2)

  out_peco <- list(fit_train=fit_train,
            fit_test=fit_test)
  saveRDS(out, file.path(paste0("data/ourdata_peco_",
                                ind, "_top", sprintf("%03d", ngenes),"genes.rds")))
}


# compute prediction error
inds <- c("NA19098","NA18511","NA18870","NA19101","NA18855","NA19160")
ngenes <- 5
diff_peco <- do.call(rbind, lapply(1:length(inds), function(i) {
  ind <- inds[i]
  fl_name <- list.files(file.path("data"),
                        pattern=paste0("ourdata_peco_", ind, "_top", sprintf("%03d", ngenes),"genes.rds"),
                        full.names = TRUE)
  res_pred <- readRDS(fl_name)$fit_test
  data_test <- readRDS(paste0("data/data_training_test/ind_",ind, "_data_test.rds"))
  all.equal(rownames(data_test$pdata_test), names(res_pred$cell_times_est))

  df_pred <- readRDS(fl_name)$fit_test
  phase_ref = data_test$theta_test
  phase_pred_rot = rotation(ref_var=data_test$theta_test, 
                          shift_var=res_pred$cell_times_est)

  return(data.frame(phase_ref = phase_ref,
                    phase_pred_rot = phase_pred_rot,
                    diff_time = circ_dist(phase_ref, phase_pred_rot),
                    ind = as.character(ind),
                    method = "peco"))
}))
saveRDS(diff_peco, file=("data/fit_diff_peco.rds"))

diff_peco %>% group_by(ind) %>% summarise(mn=mean(diff_time)/2/pi,
                                          sd=sd(diff_time/2/pi)/sqrt(sum(phase_pred_rot>0)))
```


## Seurat

We computed Seurat cell cycle scores and transformed these two scores into continuous cell cycle phase. Specifically, we applied PCA to the two cell cycle scores and then assigend each cell an angle on a unit circle. We did this for samples of each individual cell lines separately. 

```{r, eval = FALSE}
# use Seurat to assign phase for each individual
# and then compute predictione error for samples in individual cell lines
diff_seurat <- do.call(rbind, lapply(1:length(inds), function(i) {
  source("code/run_seurat.R")
  seurat.genes <- readLines(con = "data/regev_lab_cell_cycle_genes.txt")
  seurat.genes <- list(s.genes=seurat.genes[1:43],
                       g2m.genes=seurat.genes[44:97])

  #  print(i)
  data <- readRDS(paste0("data/data_training_test/ind_",inds[i], "_data_test.rds"))
  Y_fit <- data$log2cpm_test
  rownames(Y_fit) <- fdata$name[match(rownames(Y_fit), rownames(fdata))]
  fit <- run_seurat(Y_fit,
                    s.genes=seurat.genes$s.genes,
                    g2m.genes=seurat.genes$g2m.genes)
  fit <- as.list(fit)
  seurat.pca <- prcomp(cbind(fit$G2M, fit$S), scale=TRUE)
  phase_pred <- intensity2circle(seurat.pca$x, plot.it = F, method = "trig")
  names(phase_pred) <- colnames(Y_fit)
  phase_ref <- pdata$theta[match(names(phase_pred), rownames(pdata))]
  phase_pred_rot <- rotation(ref_var=phase_ref, shift_var=phase_pred)
  return(data.frame(phase_ref = phase_ref,
                    phase_pred_rot = phase_pred_rot,
                    diff_time = circ_dist(phase_ref, phase_pred_rot),
                    ind = inds[i],
                    method = "Seurat"))
}) )
saveRDS(diff_seurat, file="data/fit_diff_seurat.rds")

diff_seurat %>% group_by(ind) %>% summarise(mn=mean(diff_time)/2/pi,
                                            sd=sd(diff_time/2/pi)/sqrt(sum(phase_pred_rot>0)))
```



## Cyclone

We computed Cyclone cell cycle scores and transformed these two scores into continuous cell cycle phase. Specifically, we applied PCA to the two cell cycle scores and then assigend each cell an angle on a unit circle. We did this for samples of each individual cell lines separately. 


```{r, eval=F}
library(scran)
hs.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))

for (ind in unique(pData(eset)$chip_id)) {
   data <- readRDS(file.path(paste0("data/data_training_test/ind_",
                                    ind, "_data_test.rds")))
   input <- data.frame(data$log2cpm_quant_test)
   ii <- which(colSums(is.na(input))==0)
   input <- input[,ii]

   # --- begin cyclone
   hs.pairs <- readRDS(system.file("exdata", "human_cycle_markers.rds", package="scran"))
#   message("run cyclone", "\n")
   out <- cyclone(input, pairs = hs.pairs,
                  gene.names=rownames(input),
                  iter=1000, min.iter=100, min.pairs=50,
                  BPPARAM=SerialParam(), verbose=T, subset.row=NULL)

   out_cyclone <- data.frame(phase_cyclone = out$phases,
                             normalized.scores_G1 = out$normalized.scores$G1,
                             normalized.scores_S = out$normalized.scores$S,
                             normalized.scores_G2M = out$normalized.scores$G2M)
   rownames(out_cyclone) <- colnames(input)

   # Infer phase using Seurat
   pc_cyclone <- prcomp(out$normalized.scores, scale. = T, center = T)

   library(peco)
   phase_peco <- intensity2circle(pc_cyclone$x[,1:2], plot.it = F, method = "trig")
   out_cyclone$phase_pred <- rotation(data$theta_test,
                                          phase_peco)
   out_cyclone$phase_ref <- data$theta_test
   out_cyclone$ind <- ind

   saveRDS(out_cyclone, file = paste0("data/ourdata_cyclone_",
                                   ind,".rds")) 
}


# compute prediction error of cyclone results
inds <- c("NA19098","NA18511","NA18870","NA19101","NA18855","NA19160")
diff_cyclone <- do.call(rbind, lapply(1:length(inds), function(i) {
  out_cyclone <- readRDS(paste0("data/ourdata_cyclone_",inds[i],".rds"))
  phase_peco <- pdata$theta[match(rownames(out_cyclone), rownames(pdata))]
  phase_cyclone_rot <- rotation(ref_var= phase_peco, shift_var = out_cyclone$phase_pred)
  return(data.frame(phase_ref = phase_peco,
                    phase_pred_rot = phase_cyclone_rot,
                    diff_time = circ_dist(phase_peco, phase_cyclone_rot),
                    ind = inds[i],
                    method = "Cyclone"))
}) )
saveRDS(diff_cyclone, file="data/fit_diff_cyclone.rds")

diff_cyclone %>% group_by(ind) %>% summarise(mn=mean(diff_time)/2/pi,
                                             sd=sd(diff_time/2/pi)/sqrt(sum(phase_pred_rot>0)))
```



## Oscope

We applied Oscope to estimate cell cycle ordering across the 888 single-cell samples. The analysis used 366 genes that were selected using Oscope's criterion of high variabilty genes. Specifically, we changed the default 'MeanCutlow` option of 100 to 10. We also found that the results were similar when adding more genes in the analysis. 


```{r, eval = F}
library(Oscope)

# to run Oscope
# use gene symbols as labels 
rownames(counts) <- fdata$name

Sizes <- MedianNorm(counts)
DataNorm <- GetNormalizedMat(counts, Sizes)
MV <- CalcMV(Data = as.matrix(counts), Sizes = Sizes, MeanCutLow = 10)
DataNorm_high_var <- DataNorm[MV$GeneToUse,]
DataInput <- NormForSine(DataNorm_high_var)
dim(DataInput)

# select samples from all individuals
SineRes <- OscopeSine(DataInput, parallel=T)
KMRes <- OscopeKM(SineRes, maxK = NULL)

# to flag clusters...
# no need to flag clusters
KMResUse <- KMRes
ENIRes <- OscopeENI(KMRes = KMResUse, Data = DataInput,
		    NCThre = 100, parallel=T)

save(DataInput,
     KMResUse, ENIRes,
     file = file.path("data/ourdata_oscope_366genes.rda")



# Oscope results
load("data/ourdata_oscope_366genes.rda")
samples_ordered <- colnames(DataInput)[ENIRes[["cluster2"]]]

phase_pred <- seq(0, 2*pi, length.out= length(ENIRes[["cluster2"]]))
names(phase_pred) <- samples_ordered

diff_oscope <- data.frame(phase_ref=pdata$theta[match(samples_ordered, colnames(eset))])
diff_oscope$phase_pred_rot <- rotation(ref_var= diff_oscope$phase_ref, shift_var = phase_pred)
diff_oscope$diff_time <- circ_dist(diff_oscope$phase_ref, diff_oscope$phase_pred_rot)
diff_oscope$ind <- pdata$chip_id[match(names(phase_pred), colnames(counts))]
diff_oscope$method <- "Oscope"

saveRDS(diff_oscope, file="data/fit_diff_oscope.rds")

diff_oscope %>% group_by(ind) %>% summarise(mn=mean(diff_time)/2/pi,
                                            sd=sd(diff_time/2/pi)/sqrt(sum(phase_pred_rot>0)))
```


## recAT

We applied reCAT to estimate cell cycle ordering across the 888 single-cell samples. The analysis used all the 11,040 genes. 

To run recAT, please clone the reCAT GitHUb repository and then cd to the directory.

`git clone https://github.com/tinglab/reCAT`

`cd "reCAT/R"`


```{r, eval = FALSE}
# recat
input <- t(log2cpm_quant)
input <- input[sample(1:nrow(input)),]
source("get_test_exp.R")
test_exp <- get_test_exp(t(input))
source("get_ordIndex.R")
res_ord <- get_ordIndex(test_exp, 10)
ordIndex <- res_ord$ordIndex

save(test_exp, ordIndex,
     file = file.path("data/ourdata_recat_rda"))

# compute predictor error of reCAT predicted time
sample_ordered <- rownames(test_exp)[ordIndex]
phase_pred <- seq(0, 2*pi, length.out= length(ordIndex))
names(phase_pred) <- sample_ordered

diff_recat <- data.frame(phase_ref=pdata$theta[match(samples_ordered, colnames(eset))])
diff_recat$ind <- pdata$chip_id[match(names(phase_pred), colnames(eset))]

diff_recat$phase_pred_rot <- rotation(ref_var= diff_recat$phase_ref, shift_var = phase_pred)
rownames(diff_recat) <- samples_ordered

diff_recat$diff_time <- circ_dist(diff_recat$phase_ref, diff_recat$phase_pred_rot)
diff_recat$method <- "reCAT"
diff_recat <- diff_recat[,c(1,3,4,2,5)]

saveRDS(diff_recat, file="data/fit_diff_recat.rds")

diff_recat %>% group_by(ind) %>% summarise(mn=mean(diff_time)/2/pi,
                                           sd=sd(diff_time/2/pi)/sqrt(sum(phase_pred_rot>0)))
```



## Session information

```{r}
sessionInfo()
```

